Election 2024

### 2024-11-02 paper/paper.qmd
Paper edits

```{python}
# Define states and electoral votes, ensuring votes are listed in reverse for clarity in plot
states = ["California", "Connecticut", "Delaware", "Hawaii", "Illinois", "Maine", 
          "Maryland", "Massachusetts", "Minnesota", "New Jersey", "New York", "Oregon", 
          "Rhode Island", "Vermont", "Washington", "District of Columbia", "New Mexico", 
          "New Hampshire", "Colorado", "Virginia", "Arizona", "Georgia", "Michigan", 
          "Nevada", "North Carolina", "Pennsylvania", "Wisconsin", "Alabama", "Alaska", 
          "Arkansas", "Florida", "Idaho", "Indiana", "Iowa", "Kansas", "Kentucky", 
          "Louisiana", "Mississippi", "Missouri", "Montana", "Nebraska", "North Dakota", 
          "Ohio", "Oklahoma", "South Carolina", "South Dakota", "Tennessee", "Texas", 
          "Utah", "West Virginia", "Wyoming"]
votes = [54, 7, 3, 4, 19, 4, 10, 11, 10, 14, 28, 8, 4, 3, 12, 3, 5, 4, 10, 13, 
         11, 16, 15, 6, 16, 19, 10, 9, 3, 6, 30, 4, 11, 6, 6, 8, 8, 6, 10, 4, 5, 
         3, 17, 7, 9, 3, 11, 40, 6, 4, 3]

@@

# Threshold for filling
dem_threshold = np.argmax(cumulative_votes >= 227)
plt.fill_between(x[:dem_threshold + 1], cumulative_votes[:dem_threshold + 1], step='post', color='blue', alpha=0.4)

# Adding threshold line and markers
plt.axhline(270, color='purple', linestyle='--', linewidth=2)
plt.text(len(states) / 2, 275, '270 votes to win', horizontalalignment='center', color='purple', fontsize=12)
plt.xlabel('Electoral College by State', fontsize=12)
plt.ylabel('Total Votes', fontsize=12)
plt.title('Cumulative Electoral Votes by State', fontsize=15)
plt.grid(True, linestyle='--', alpha=0.5)
plt.tight_layout()
plt.show()

@@

```{python}
#| echo: false

# Load the data (replace with your CSV file path)
df = pd.read_csv('../data/02-analysis_data/merged_swing_state_data.csv')

@@

# Split 'variable_stat' into 'variable' and 'stat'
summary_long[['variable', 'stat']] = summary_long['variable_stat'].str.rsplit('_', n=1, expand=True)

@@

# Print the table using tabulate
print(tabulate(rows, headers=['State', 'Variable', 'Mean', 'Std', 'Min', 'Max', 'Count'], tablefmt='fancy_grid'))
```

# 2.3 Variables of Interest
In our logistic regression model predicting the 2024 U.S. Presidential Election outcome, the primary variables of interest are derived from aggregated state-level polling data. The percentage of respondant favouring a candidate become the key variables for prediction. Within the collection of polls in our sample the independent variable `trump_pct` represents the percentage of respondents who indicate support for Donald Trump , and `harris_pct` respectively.
Polls ar taking towards the build up of an election and thus ar taken by repondants at different times. Graph  x shows how the sample of polls are distributed over time in which it is clear that we have a collection of polls from a wide range of periods. Because sentiment changes over time we will adjust the weight that we place on polls that are closer to the election as these will be more indicative of how respondants will be voting on the day.
```{python}
#| echo: false

@@

plt.figure(figsize=(12, 6))
plt.hist(data['end_date'], bins=20, edgecolor='black', alpha=0.7)
plt.xlabel('Poll End Date')
plt.ylabel('Poll Count')
plt.title('Poll Distribution Over Time')
plt.xticks(rotation=45)
plt.tight_layout()
plt.show()

@@

 ```{Python}
#| echo: false

import pandas as pd
from tabulate import tabulate

# Data setup
data = {
    'State': ['Arizona', 'Georgia', 'Michigan', 'Nevada', 'North Carolina', 'Pennsylvania', 'Wisconsin'],
    'Trump % (Mean)': [49.85, 49.39, 48.57, 47.9, 48.48, 47.68, 47.76],
    'Trump % (SD)': [1.24, 1.78, 2.14, 1.5, 1.41, 1.81, 2.0],
    'Harris % (Mean)': [47.04, 47.66, 47.72, 48.63, 48.58, 48.53, 48.92],
    'Harris % (SD)': [2.17, 2.52, 2.12, 1.19, 1.55, 1.29, 1.72],
    'Percentage Chance of Trump Winning': [86.85, 71.18, 53.09, 21.34, 32.31, 19.67, 15.52]
}

@@

# Display as a professional table
table_output = tabulate(
    df, headers='keys', tablefmt='grid', showindex=False,
    floatfmt=(".2f", ".2f", ".2f", ".2f", ".2f", ".2f")
)
print(table_output)

in figure xx we cans see th results from the model described.

@@

```{python}
#| echo: false

# Data setup (can adjust based on model results)
data = {
    'State': ['Arizona', 'Georgia', 'Michigan', 'Nevada', 'North Carolina', 'Pennsylvania', 'Wisconsin'],
    'Trump % (Mean)': [49.85, 49.39, 48.57, 47.9, 48.48, 47.68, 47.76],
    'Trump % (SD)': [1.24, 1.78, 2.14, 1.5, 1.41, 1.81, 2.0],
    'Harris % (Mean)': [47.04, 47.66, 47.72, 48.63, 48.58, 48.53, 48.92],
    'Harris % (SD)': [2.17, 2.52, 2.12, 1.19, 1.55, 1.29, 1.72],
    'Trump Win Probability (%)': [86.85, 71.18, 53.09, 21.34, 32.31, 19.67, 15.52]
}
df = pd.DataFrame(data)
df['Distance from 50%'] = df['Trump Win Probability (%)'] - 50

# Centered bar chart
fig, ax = plt.subplots(figsize=(10, 6))
colors = ['red' if x > 0 else 'blue' for x in df['Distance from 50%']]
ax.barh(df['State'], df['Distance from 50%'], color=colors)
ax.axvline(0, color='black')
ax.set_xlabel('Distance from 50% (Trump Winning Probability)')
ax.set_title('Probability of Trump Win, Centered from 50%')
plt.tight_layout()
plt.show()

@@

While our model provides a structured framework for predicting the election, it has several limitations. First, the model is based on polling data available far in advance of the election, meaning there is a high degree of uncertainty, and the model may not fully capture late shifts in public opinion or external shocks (e.g., economic downturns or scandals). Additionally, the use of linear modeling may oversimplify the complex dynamics of voter behavior, as elections often involve non-linear influences that are difficult to predict.

@@

>>>>

```{python}
#| ccho: false
#| fig-cap: "This figure shows the cumulative total of electoral votes by state, grouped by current party affiliation, accordng to 538. The chart distinguishes between Democrat, Republican, and swing states, with a color-coded fill under each party’s cumulative curve. Key threshold of 270 electoral vote mark needed to secure a victory is diplayed. This visualization shows the democrats lead with 227 votes and the Republicans with 219 vot secured"
# Define states and electoral votes, ensuring votes are listed in reverse for clarity in plot

states = ["California", "Connecticut", "Delaware", "Hawaii", "Illinois", "Maine", 
          "Maryland", "Massachusetts", "Minnesota", "New Jersey", "New York", "Oregon", 
          "Rhode Island", "Vermont", "Washington", "District of Columbia", "New Mexico", 
          "New Hampshire", "Colorado", "Virginia", "Arizona", "Georgia", "Michigan", 
          "Nevada", "North Carolina", "Pennsylvania", "Wisconsin", "Alabama", "Alaska", 
          "Arkansas", "Florida", "Idaho", "Indiana", "Iowa", "Kansas", "Kentucky", 
          "Louisiana", "Mississippi", "Missouri", "Montana", "Nebraska", "North Dakota", 
          "Ohio", "Oklahoma", "South Carolina", "South Dakota", "Tennessee", "Texas", 
          "Utah", "West Virginia", "Wyoming"]
votes = [54, 7, 3, 4, 19, 4, 10, 11, 10, 14, 28, 8, 4, 3, 12, 3, 5, 4, 10, 13, 
         11, 16, 15, 6, 16, 19, 10, 9, 3, 6, 30, 4, 11, 6, 6, 8, 8, 6, 10, 4, 5, 
         3, 17, 7, 9, 3, 11, 40, 6, 4, 3]

@@

# Threshold for filling
dem_threshold = np.argmax(cumulative_votes >= 227)
plt.fill_between(x[:dem_threshold + 1], cumulative_votes[:dem_threshold + 1], step='post', color='blue', alpha=0.4)
threshold_index = np.argmax(cumulative_votes >= 219)
plt.fill_between(x_reversed[:threshold_index + 5], cumulative_votes_reversed[:threshold_index + 5], step='post', color='red', alpha=0.4)

# Adding threshold line and markers
plt.axhline(270, color='purple', linestyle='--', linewidth=2)
plt.text(len(states) / 2, 275, '270 votes to win', horizontalalignment='center', color='purple', fontsize=12)
plt.xlabel('Electoral College by State', fontsize=12)
plt.ylabel('Total Votes', fontsize=12)
plt.title('Cumulative Electoral Votes by State', fontsize=25)
plt.grid(True, linestyle='--', alpha=0.5)
plt.tight_layout()
plt.show()

@@

```{python}

#| echo: false
#| fig-cap: "Summary statistics of key sample variabls by state. Shows high sampl size and numeric grade among all polls of swing states."
# Load the data (replace with your CSV file path)
df = pd.read_csv('../data/02-analysis_data/merged_swing_state_data.csv')

@@

# Rename elements
summary_long[['variable', 'stat']] = summary_long['variable_stat'].str.rsplit('_', n=1, expand=True)
summary_long['variable'] = summary_long['variable'].str.replace('_', ' ', regex=False).str.title()

@@

# Print the table using tabulate
print(tabulate(rows, headers=['State', 'Variable', 'Mean', 'Std', 'Min', 'Max', 'Count'], tablefmt='simple_grid'))
```
**Table 1: Summary statistics of key sample variabls by state. Shows high sampl size and numeric grade among all polls of swing states.**

# 2.3 Variables of Interest
In our logistic regression model predicting the 2024 U.S. Presidential Election outcome, the primary variables of interest are derived from aggregated state-level polling data. The percentage of respondant favouring a candidate become the key variables for prediction. Within the collection of polls in our sample the independent variable `trump_pct` represents the percentage of respondents who indicate support for Donald Trump , and `harris_pct` respectively.
Polls ar taking towards the build up of an election and thus ar taken by repondants at different times. Graph  x shows how the sample of polls are distributed over time in which it is clear that we have a collection of polls from a wide range of periods. Because sentiment changes over time we will adjust the weight that we place on polls that are closer to the election as these will be more indicative of how respondants will be voting on the day.
```{python}
#| echo: false
#| fig-cap: "A histogram of the poll quantities, aggregated by the end date of each poll, shows the distribution of polling data over time. This chart reveals the frequency of polls leading up to the election and provides insight into polling activity trends, with potential spikes reflecting periods of increased public interest or campaign activity."

@@

plt.figure(figsize=(12, 6))
plt.hist(data['end_date'], bins=20, edgecolor='black', alpha=0.7)
plt.xlabel('Poll End Date')
plt.ylabel('Poll Count')
plt.title('Poll Distribution Over Time', fontsize=25)
plt.xticks(rotation=45)
plt.tight_layout()
plt.show()

@@

```{python}
#| echo: false
# Data setup
data = {
    'State': ['Arizona', 'Georgia', 'Michigan', 'Nevada', 'North Carolina', 'Pennsylvania', 'Wisconsin'],
    'Trump %': [49.85, 49.39, 48.57, 47.9, 48.48, 47.68, 47.76],
    'Trump SD': [1.24, 1.78, 2.14, 1.5, 1.41, 1.81, 2.0],
    'Harris %': [47.04, 47.66, 47.72, 48.63, 48.58, 48.53, 48.92],
    'Harris SD': [2.17, 2.52, 2.12, 1.19, 1.55, 1.29, 1.72],
    'Trump Winning %': [86.85, 71.18, 53.09, 21.34, 32.31, 19.67, 15.52]
}

@@

# Display as a professional table
print(tabulate(
    df, headers='keys', tablefmt='grid', showindex=False,
    floatfmt=(".2f", ".2f", ".2f", ".2f", ".2f", ".2f")
))

```
**Table 2: Results table for the model describd above. Show support fo trump and Kamala at the stat level accordng to avalable polls. Small standard dviation show strong results, yet the rac is very close in many states. Percentages above 50% show the favoured candidate in each state.**

in figure xx we cans see the results from the model described.

@@

```{python}
#| echo: false
#| fig-cap: "A centered bar chart where each state has a bar reflecting Trump’s probability of winning. The bars extend to the right or left from the center line at 50%, with the distance representing the magnitude of deviation from an even chance. This chart visualizes Trump’s relative strength in each swing state and highlights that he s only favoured in Georgia and Arizona."
# Data setup (can adjust based on model results)
data = {
    'State': ['Arizona', 'Georgia', 'Michigan', 'Nevada', 'North Carolina', 'Pennsylvania', 'Wisconsin'],
    'Trump % (Mean)': [49.85, 49.39, 48.57, 47.9, 48.48, 47.68, 47.76],
    'Trump % (SD)': [1.24, 1.78, 2.14, 1.5, 1.41, 1.81, 2.0],
    'Harris % (Mean)': [47.04, 47.66, 47.72, 48.63, 48.58, 48.53, 48.92],
    'Harris % (SD)': [2.17, 2.52, 2.12, 1.19, 1.55, 1.29, 1.72],
    'Trump Win Probability (%)': [86.85, 71.18, 53.09, 21.34, 32.31, 19.67, 15.52]
}
df = pd.DataFrame(data)
df['Distance from 50%'] = df['Trump Win Probability (%)'] - 50

# Centered bar chart
fig, ax = plt.subplots(figsize=(10, 6))
colors = ['red' if x > 0 else 'blue' for x in df['Distance from 50%']]
ax.barh(df['State'], df['Distance from 50%'], color=colors)
ax.axvline(0, color='black')
ax.set_xlabel('Distance from 50% (Trump Winning Probability)')
ax.set_title('Probability of Trump Win in swing states, Centered from 50%', fontsize=25)
plt.tight_layout()
plt.show()

@@

While our model provides a structured framework for predicting the election, it has several limitations. First, the model is based on polling data available far in advance of the election, meaning there is a high degree of uncertainty, and the model may not fully capture late shifts in public opinion or external shocks (e.g., economic downturns or scandals). As polling results are based on public opinion, large shock events are likely and expected, moreover scheduled events such as debates and rallies can be expected to have large impacts on the election that cannot be modeled here. Additionally, the use of linear modeling may oversimplify the complex dynamics of voter behavior, as elections often involve nonlinear influences that are difficult to predict.

@@

Candidate Trump’s low likelihood of winning shown by our results may under-represent his support, this claim is supported by the disparity with the betting markets. This could be a result of a common problem with polling data, response rate bias. In the previous two elections, where Donald Trump was also the republican candidate, his polling was far below the eventual results on election day. Post-mortems of both elections indicated evidence of those supporting Trump not participating in polls and thus under-presenting Republican support Clinton t al (2021) kennedy et al (2018)]. Should this trend continue, and be present in the polls used in this analysis, it may be expected that Trump could surprise the model and win some of these swing states. Another reason Trump could outperform the model is another bias often seen in surveys, one of social desirability. Trump is a unique candidate as, if elected, he would become the first president to be elected with federal crime conviction. This along with other negative publicity the former host of tv’s apprentice has garnered since the start of his political campaign could bias voters from publicly or privately showing their support. Even with the quality polls used in this analysis, where steps are taken to ensure representative results(see appendix B), we could be observing a skew in the results.
